HMM

### 2015-12-26 paper3.Rnw

\begin{abstract}
Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal heterogeneity and
autocorrelation (persistence) in group occupancy can obscure such
signals. We use latent-state and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers. Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely explored model extension, resolves previous HMM modeling issues
and clarifies the behavioural patterns of panthers.
...\end{abstract}

\keywords{Hidden Markov, temporal heterogeneity}

@@

Given a sequence of animal movements observed via direct observation
or telemetry, movement models aim to find a parsimonious description
that can be used to understand past movements as well as predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012}
-- formalized in ecological terms as the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} -- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behavior is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin1998quantitative}.

@@

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states stays
relatively constant; changes in behavioural mode switching based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred state and environmental
conditions \citep{fryxell_multiple_2008}.\cite{schliehe-diecks_application_2012}
consider changes in switching behaviour over time scales of six-hour
observation periods, but
in general ecologists have used other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are independent,
\emph{finite mixture models} (FMM) -- a special case of HMMs where the
probability of state occupancy is independent of the previous state --
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities changed deterministically over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or switching
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying switching probabilities -- in particular,
switching probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We take data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models, HSMMs (an extension of HMMs that allows flexible modelling of dwell time)

@@

\bibliography{paper.bib}

>>>>

\begin{abstract}

Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal heterogeneity and
autocorrelation (persistence) in group occupancy can obscure such
signals. We use finite mixture and hidden Markov models, two standard
clustering techniques, to model high-resolution hourly movement data
from Florida panthers. Allowing for temporal heterogeneity in
transition probabilities, a straightforward but rarely explored model
extension, resolves some shortcomings of current modeling frameworks
and clarifies the behavioural patterns of panthers.

\end{abstract}

\keywords{hidden Markov model, animal movement, temporal autocorrelation, temporal heterogeneity}

@@

Given a sequence of animal movements observed directly 
or via telemetry, movement models aim to find a parsimonious description
that can be used to understand past movements as well as predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} %
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behavior is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin1998quantitative}.

@@

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states is
approximately constant. Changes in behavioural mode switching 
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in switching behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are effectively independent,
\emph{finite mixture models} (FMM) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or switching
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying switching probabilities -- in particular,
switching probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We re-analyze data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
analyze the movement of Florida panthers.

@@

\bibliography{paper}
